---
title: "Bladen County Absentee Ballots in 2018 Election"
author: "Matt DeVries"
date: "12/4/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Map and Tables of US Congress 9th District for N.C. Race

Bladen County, North Carolina may have been a target of a campaign of electoral fraud, according to media reports <https://www.nytimes.com/2018/12/04/us/politics/north-carolina-election-fraud.html>. 
Dr. Michael Bitzer of Catawba College has identified several unusual trends evident in the race for US Congress in the 9th District <http://www.oldnorthstatepolitics.com/2018/12/NCs-9th-competitive-and-contested-CD.html>.  Dr. Bitzer reports unusually low rates of acceptance for absentee mail-in ballots for voters in Bladen County who voted in the US Congress 9th District race. When this report was published, Republican Mark Harris led his opponent Democrat Dan McCready by less than 1,000 votes in this competitive contest. 

This analysis attempts to look deeper into Bladen County by generating a map of voters, which takes into account voters' race. Using an interactive map, this analysis identifies areas of concern where voters may have been targeted.  This map allows a geographic overview of where rejected ballots were cast by African-American voters.

## Methods
This analysis was generated using the R programming language using data from the North Carolina State Board of Elections (NCSBE).  Data on election results and voters are available online for download <https://dl.ncsbe.gov/index.html>.
```{r warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(knitr)
library(DT)
library(leaflet)
library(scales)
```


NCSBE makes available an 800MB file with details on all votes cast in this election, including all requests for absentee ballots.  Because of the size of this data file and its' public availability from NCSBE, this file isn't stored in the github repository for this analysis.  The file absentee_20181106.csv was downloaded for this report on December 4, 2018.

```{r eval=FALSE}
all_absentee <- read.csv("~/Desktop/NC2018/absentee_20181106.csv")

## Filter Mail-In, US Congress 9 Race, Bladen County
allData9thBladenMail <- all_absentee %>% 
        filter(cong_dist_desc == "CONGRESSIONAL DISTRICT 9" & county_desc == "BLADEN") %>% 
        filter(ballot_req_delivery_type == "MAIL")

## CREATE ADDRESS FOR GEOCODING
allData9thBladenMail <- allData9thBladenMail %>% 
        mutate(fullAddress = paste0(voter_street_address, ", ", voter_city, ", NC ", voter_zip))

## Write pre-geocoded data to file
write.csv(allData9thBladenMail, file = "geocode/Mail9thBladenx.csv")
```

The analysis relies on the geocoding service available at geocod.io.  Addresses of voters who attempted to vote with mail-in ballots were geocoded, generating latitude and longitude locations for the voters.  For this analysis, a copy of the ballots data was read from a file downloaded from geocodd.io.

```{r}
## Load data analyzed with geocodio
Mail9thBladen <- read.csv("geocode/Mail9thBladenx_geocodio.csv")
names(Mail9thBladen)[names(Mail9thBladen)=="fullAdress"] <- "fullAddress"
```


Additional data about voters was retrieved from the NCSBE about voter's ethnicity in Bladen County.  Additionally, press reports identified several voters who suspected that they had been targeted as part of a campaign of electoral fraud.  These voters location information was retrieved in order to identify potential areas where fraud occurred.
```{r}
## Voters in Bladen County, includes ethnicity Data
bladenVoters <- read.table("data/ncvoter9.txt", header = T)

bladenVoters <- bladenVoters %>% 
        select(county_desc, voter_reg_num, last_name, first_name, middle_name,
               name_suffix_lbl, res_street_address, res_city_desc, state_cd,
               zip_code, race_code, ethnic_code, party_cd, gender_code,
               birth_age, precinct_abbrv, precinct_desc, cong_dist_abbrv)

## Voters mentioned in press
votersPress <-  Mail9thBladen %>% 
  filter(voter_first_name %in% "EMMA" & voter_last_name %in% "SHIPMAN") 
votersPress <-  Mail9thBladen %>% 
  filter(voter_first_name %in% "DATESHA" & voter_last_name %in% "MONTGOMERY") %>% 
  bind_rows(votersPress)
votersPress <-  Mail9thBladen %>% 
  filter(voter_first_name %in% "JENEVA" & voter_last_name %in% "LEGIONS") %>% 
  bind_rows(votersPress)

suspectAreas <- votersPress %>% 
  filter(ballot_rtn_status != "ACCEPTED")

## Add race2 variable with Latino to Mail9thBladen
bladenEthnic <- bladenVoters %>% 
        select(voter_reg_num,
        ethnic_code)

Mail9thBladen <- left_join(x = Mail9thBladen, y = bladenEthnic, by = c("voter_reg_num", "voter_reg_num"))

Mail9thBladen <- Mail9thBladen %>% 
        mutate(race2 = case_when(
                (race=="WHITE" & ethnic_code != "HL") ~ "White",
                race=="BLACK or AFRICAN AMERICAN" ~ "African-American",
                race=="WHITE" & ethnic_code == "HL" ~ "Latino",
                race=="OTHER" ~ "Other",
                race == "UNDESIGNATED" ~ "Undesignated",
                race== "INDIAN AMERICAN or ALASKA NATIVE" ~ "Native Am."
        ))

Mail9thBladen$race2 <- as.factor(Mail9thBladen$race2)
```

## Map of Mail-In Ballots
This analysis includes a map generated from the addresses of voters who tried to vote via mail-in ballot in the race for US Congress 9th District.  Voters are characterized as being either African-American or another race other than African-American, including white voters.  Voters are further categorized as either having their mail-in ballot accepted or rejected.  Rejected ballots include both ballots that were received by the elections board but not counted and ballots that were never received by the election board.

```{r main map}
## Add Variaable coming race and ballot status
Mail9thBladen <-  Mail9thBladen %>% 
  mutate(mapRaceBallot = ifelse(race2=="African-American", "African-American ", "Other Race ")) %>% 
  mutate(mapRaceBallot = paste0(mapRaceBallot, ifelse(ballot_rtn_status=="ACCEPTED", "Accepted", "Rejected")))

Mail9thBladen$mapRaceBallot <- as.factor(Mail9thBladen$mapRaceBallot)

levelsRace3 <- levels(Mail9thBladen$mapRaceBallot)

cofrace3 <- colorFactor(c("#339933", "#FF3030", "#DDDD00", "#888888"), 
                        domain=levelsRace3)
```

Below is the map of all mail-in ballots for the 9th Congressional District, categorized by race and whether the ballot was accepted. The larger gray circles are around areas where press reports indicate that voters may have been targeted as part of a campaign of voter fraud. 
```{r}
m <- leaflet(Mail9thBladen) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  setView(-78.604867, 34.629176, zoom = 10) %>% 
  addCircles(suspectAreas$Longitude, suspectAreas$Latitude,
             radius = 2000, color = "#444444", stroke = T, fillOpacity = 0.1
  ) %>% 
  addCircles(~Longitude, ~Latitude, popup= ~fullAddress,
             weight = 3, radius=40, 
             color=~cofrace3(mapRaceBallot), stroke = TRUE, fillOpacity = 0.3) %>% 
  addLegend(colors = c("#339933", "#FF3030", "#DDDD00", "#888888"), 
            labels = c(levelsRace3))
m
```

## Additional Analysis of Ballots and Voters

This analysis includes a comparison between the rate of acceptance of ballots cast by African-American voters and white voters.  The two largest cities in the Bladen county involved in this contest are Bladenboro and Elizabethtown.  Ballots cast by African-Americans in Baldenboro were accepted at an unusually low rate, less than 20%.  Below is a tabulation of the votes for this part of the analysis.

```{r}
tableCity2 <- Mail9thBladen %>% 
        group_by(voter_city, race2) %>% 
        count(ballot_rtn_status) %>% 
        spread(key = ballot_rtn_status, value = n)

## Turns NA's to Zero's
tableCity2$ACCEPTED <- ifelse(is.na(tableCity2$ACCEPTED), 0, tableCity2$ACCEPTED)
tableCity2$V1 <- ifelse(is.na(tableCity2$V1), 0, tableCity2$V1)
tableCity2$`RETURNED UNDELIVERABLE` <- ifelse(is.na(tableCity2$`RETURNED UNDELIVERABLE`), 0, tableCity2$`RETURNED UNDELIVERABLE`)
tableCity2$SPOILED <- ifelse(is.na(tableCity2$SPOILED), 0, tableCity2$SPOILED)
tableCity2$`VOTER SIGNATURE MISSING` <- ifelse(is.na(tableCity2$`VOTER SIGNATURE MISSING`), 0, tableCity2$`VOTER SIGNATURE MISSING`)
tableCity2$`WITNESS INFO INCOMPLETE` <- ifelse(is.na(tableCity2$`WITNESS INFO INCOMPLETE`), 0, tableCity2$`WITNESS INFO INCOMPLETE`)

## Calculate all 9th Mail-In Votes and add back to tableCity2
countyNums <- tableCity2 %>% 
        group_by(race2)  %>% 
         summarise_at(vars(V1:`WITNESS INFO INCOMPLETE`), sum, na.rm =T) 

countyNums <- countyNums %>% 
        mutate(propBallot = ACCEPTED/(V1 + ACCEPTED + `RETURNED UNDELIVERABLE` + 
             SPOILED + `VOTER SIGNATURE MISSING` + `WITNESS INFO INCOMPLETE`))
countyNums<- countyNums %>% 
        select(race2, propBallot)


countyResults <- select(spread(countyNums, key=race2, value = propBallot), `African-American`, White)
countyResults <- cbind(as.data.frame("All Ballots for US 9th"), countyResults)

names(countyResults)[names(countyResults)=="\"All Ballots for US 9th\""] <- 'Location'


tableCity2 <- tableCity2 %>% 
        mutate(propBallot = ACCEPTED/(V1 + ACCEPTED + `RETURNED UNDELIVERABLE` + 
          SPOILED + `VOTER SIGNATURE MISSING` + `WITNESS INFO INCOMPLETE`))

tableCity3 <- tableCity2 %>% 
        select( voter_city, race2, propBallot) %>% 
        filter(voter_city %in% c("BLADENBORO", "ELIZABETHTOWN"))

cityResults <- select(spread(tableCity3, key=race2, value = propBallot), 
                      voter_city, `African-American`, White)

names(cityResults)[names(cityResults)=="voter_city"] <- 'Location'

cityResults$Location <- as.character(cityResults$Location)
countyResults$Location <- as.character(countyResults$Location)

## Convert numbers to percentage
resultsTable <- bind_rows(as.data.frame(cityResults), countyResults)
resultsTable$`African-American` <- percent(resultsTable$`African-American`)
resultsTable$White <- percent(resultsTable$White)
```


## Table: Rate of Acceptance for Bladen Absentee Ballots
The table below reports the rate at which mail-in ballots for the 9th District were accepted for African-American and White Voters in two cities and the entire county. Rejected ballots include both ballots that were received by the elections board but not counted and ballots that were never received by the election board.
```{r echo=FALSE}
datatable(resultsTable)

```

## Table: All Absentee Ballots
Below is table of all mail-in ballots in Bladen County for the US 9th District.  This table can be sorted and searched.  Points in the map above have popups that reveal the street address of voters.

```{r ballots}
## Ballots for Display
ballots4display <- Mail9thBladen %>% 
  select(race2, voter_street_address, voter_city, voter_zip, voter_party_code, ballot_req_dt,
         ballot_rtn_dt, ballot_rtn_status, fullAddress)
```

```{r echo=FALSE}
datatable(ballots4display)
```
This R markdown document and some data files used in this analysis are available on GitHub under the link2matt account <https://github.com/link2matt/bladen-co-nc-election-2018>